---
title: "Lab 5"
subtitle: "Introduction to Statistical Learning - PISE"
author: "[Aldo Solari]{.orange}"
institute: "_Ca' Foscari University of Venice_"
page-layout: full
execute:
  cache: false
  error: true
filters:
  - remove-pause.lua
format:
  revealjs:
    auto-stretch: true
    center: true
    html-math-method: katex
    transition: none
    output-file: un_lab5_slides.html
    slide-number: true
    callout-appearance: minimal
    code-line-numbers: true
    theme: [default, ../template.css]
    embed-resources: false
    echo: false
    fig-dpi: 200
    logo: img/cf_logo.png
    footer: "[ISL](https://aldosolari.github.io/ISL)"
    highlight-style: github
  html:
    html-math-method: katex
    echo: false
    callout-appearance: minimal
    theme: [simplex, ../template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: true
    smooth-scroll: true
    code-fold: false
    code-summary: "Show the code"
    fig-dpi: 150
    highlight-style: github
editor:
  markdown:
    wrap: 72
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false
knitr::opts_chunk$set(error = TRUE)
```

# Unsupervised Learning

## Principal Components Analysis

- PCA is performed on the `USArrests` dataset (included in base R)  
- Rows represent the 50 U.S. states  
- States are listed in alphabetical order

```{r}
#| label: chunk1
#| echo: true
states <- row.names(USArrests)
states
```

- Columns of the dataset correspond to the four variables

```{r}
#| label: chunk2
#| echo: true
names(USArrests)
```

- Initial data exploration is performed  
- Variables have very different mean values

```{r}
#| label: chunk3
#| echo: true
apply(USArrests, 2, mean)
```

- `apply()` applies a function (e.g., `mean()`) to rows or columns of a dataset  
- Second argument specifies dimension (`1` = rows, `2` = columns)  
- Data shows large scale differences: rapes ≈ 3× murders, assaults ≈ 8× rapes  
- `apply()` can also be used to compute variances of variables


```{r}
#| label: chunk4
#| echo: true
apply(USArrests, 2, var)
```

- Variables have very different variances  
- `UrbanPop` (percentage) is not comparable to crime rates (per 100,000 people)  
- Without scaling, PCA would be dominated by `Assault` (largest mean and variance)  
- Therefore, variables must be standardized (mean = 0, sd = 1) before PCA  
- PCA is performed using the `prcomp()` function in R

```{r}
#| label: chunk5
#| echo: true
pr.out <- prcomp(USArrests, scale = TRUE)
```

- `prcomp()` centers variables to mean 0 by default  
- `scale = TRUE` standardizes variables to standard deviation 1  
- Output of `prcomp()` includes several useful components

```{r}
#| label: chunk6
#| echo: true
names(pr.out)
```

- `center`: means of variables used for centering  
- `scale`: standard deviations used for scaling before PCA

```{r}
#| label: chunk7
#| echo: true
pr.out$center
pr.out$scale
```

- `rotation` matrix contains the principal component loadings  
- Each column represents a principal component loading vector  
- Multiplying the data matrix `X` by `rotation` gives the principal component scores  
- Called "rotation" because it represents a rotation of the original coordinate system

```{r}
#| label: chunk8
#| echo: true
pr.out$rotation
```

- There are 4 principal components  
- In general, number of informative PCs = $min(n − 1, p)$ 
- `prcomp()` directly provides principal component scores  
- `pr.out$x` is a 50 × 4 matrix of scores  
- Each column corresponds to a principal component score vector

```{r}
#| label: chunk9
#| echo: true
dim(pr.out$x)
```

- The first two principal components can be visualized using a plot

```{r}
#| label: chunk10
#| echo: true
biplot(pr.out, scale = 0)
```

- In `biplot()`, `scale = 0` scales arrows to represent loadings  
- Different `scale` values change interpretation of the biplot  
- PCA results are only unique up to a sign change  
- Plots may appear as mirror images due to sign ambiguity  
- Sign flipping can reproduce equivalent visualizations

```{r}
#| label: chunk11
#| echo: true
pr.out$rotation = -pr.out$rotation
pr.out$x = -pr.out$x
biplot(pr.out, scale = 0)
```

- `prcomp()` outputs the standard deviation of each principal component  
- These can be accessed via `pr.out$sdev`

```{r}
#| label: chunk12
#| echo: true
pr.out$sdev
```

- Variance explained by each principal component is obtained by squaring the standard deviations  
- Computed as: `pr.var <- pr.out$sdev^2`

```{r}
#| label: chunk13
#| echo: true
pr.var <- pr.out$sdev^2
pr.var
```

- Proportion of variance explained (PVE) is computed by dividing each component’s variance by total variance  
- Computed as: `pve <- pr.var / sum(pr.var)`

```{r}
#| label: chunk14
#| echo: true
pve <- pr.var / sum(pr.var)
pve
```

- First principal component explains ~62.0% of the variance  
- Second principal component explains ~24.7%  
- Remaining components explain smaller proportions  
- PVE and cumulative PVE can be visualized with plots

```{r}
#| label: chunk15
#| echo: true
par(mfrow = c(1, 2))
plot(pve, xlab = "Principal Component",
    ylab = "Proportion of Variance Explained", ylim = c(0, 1),
    type = "b")
plot(cumsum(pve), xlab = "Principal Component",
    ylab = "Cumulative Proportion of Variance Explained",
    ylim = c(0, 1), type = "b")
```

- Results are visualized in a plot (Figure 12.3)  
- `cumsum()` computes the cumulative sum of a numeric vector  
- Used to calculate cumulative proportion of variance explained

```{r}
#| label: chunk16
#| echo: true
a <- c(1, 2, 8, -3)
cumsum(a)
```

## Matrix Completion

- Recreate analysis on `USArrests` dataset  
- Convert data frame to matrix  
- Center and scale each column (mean = 0, variance = 1)

```{r}
#| label: chunk17
#| echo: true
X <- data.matrix(scale(USArrests))
pcob <- prcomp(X)
summary(pcob)
```

- First principal component explains ~62% of the variance  
- PCA can be formulated as an optimization problem on centered data  
- Computing the first $M$ principal components solves this problem  
- Singular Value Decomposition (SVD) is a general method to compute PCA

```{r}
#| label: chunk18
#| echo: true
sX <- svd(X)
names(sX)
round(sX$v, 3)
```

- `svd()` returns three components: `u`, `d`, and `v`  
- Matrix `v` corresponds to PCA loadings (up to a sign flip)

```{r}
#| label: chunk19
#| echo: true
pcob$rotation
```

- Matrix `u` contains standardized principal component scores  
- Vector `d` contains singular values (related to standard deviations)  
- PCA scores can be recovered from SVD output  
- These scores are identical to those from `prcomp()`

```{r}
#| label: chunk20
#| echo: true
t(sX$d * t(sX$u))
pcob$x
```

- Although `prcomp()` could be used, `svd()` is used here for illustration  
- Randomly remove 20 entries from the 50 × 4 data matrix  
- Procedure:
  - Select 20 rows (states) at random  
  - For each selected row, randomly remove one of the four variables  
- Ensures each row still has at least three observed values

```{r}
#| label: chunk21
#| echo: true
nomit <- 20
set.seed(15)
ina <- sample(seq(50), nomit)
inb <- sample(1:4, nomit, replace = TRUE)
Xna <- X
index.na <- cbind(ina, inb)
Xna[index.na] <- NA
```

- `ina`: indices of rows (states) with missing values  
- `inb`: indices of columns (features) with missing values  
- `index.na`: matrix combining row and column indices for missing entries  
- Demonstrates indexing a matrix using a matrix of indices  

- Implement Algorithm 12.1 for matrix completion  
- Define a function to approximate the matrix using `svd()`  
- This function is used in Step 2 of the algorithm  
- `svd()` is used instead of `prcomp()` for illustration purposes

```{r}
#| label: chunk22
#| echo: true
fit.svd <- function(X, M = 1) {
   svdob <- svd(X)
   with(svdob,
       u[, 1:M, drop = FALSE] %*%
       (d[1:M] * t(v[, 1:M, drop = FALSE]))
     )
}
```

- Explicit `return()` is not needed in R; last computed value is returned automatically  
- `with()` is used to simplify access to elements of `svdob`  
- Alternative: directly reference components (e.g., `svdob$u`, `svdob$d`, `svdob$v`) inside the function

```{r}
#| label: chunk23
#| echo: true
#| eval: false
svdob$u[, 1:M, drop = FALSE] %*%
  (svdob$d[1:M] * t(svdob$v[, 1:M, drop = FALSE]))
```

- Step 1: initialize `Xhat` (≈ $\tilde{\mathbf{X}}$)  
- Replace missing values with column means (computed from non-missing entries)

```{r}
#| label: chunk24
#| echo: true
Xhat <- Xna
xbar <- colMeans(Xna, na.rm = TRUE)
Xhat[index.na] <- xbar[inb]
```

- Prepare variables to track progress of the iterative algorithm  
- Initialize error measures and convergence threshold  
- Set up structures to monitor changes across iterations

```{r}
#| label: chunk25
#| echo: true
thresh <- 1e-7
rel_err <- 1
iter <- 0
ismiss <- is.na(Xna)
mssold <- mean((scale(Xna, xbar, FALSE)[!ismiss])^2)
mss0 <- mean(Xna[!ismiss]^2)
```

- `ismiss`: logical matrix indicating missing entries (`TRUE` = missing)  
- Allows separation of missing vs non-missing values  

- Error tracking:
  - `mss0`: mean squared value of non-missing entries  
  - `mssold`: previous mean squared error  
  - `mss`: current mean squared error  

- Convergence criterion:
  - Relative error = `(mssold - mss) / mss0`  
  - Stop when relative error < `1e-7`  
  - Scaling ensures convergence is independent of data magnitude  

- Step 2 (iterative loop):
  - (a) Approximate matrix using `fit.svd()` → `Xapp`  
  - (b) Update missing values in `Xhat` using `Xapp`  
  - (c) Compute new error and relative error  
- Implemented using a `while()` loop

```{r}
#| label: chunk26
#| echo: true
while(rel_err > thresh) {
    iter <- iter + 1
    # Step 2(a)
    Xapp <- fit.svd(Xhat, M = 1)
    # Step 2(b)
    Xhat[ismiss] <- Xapp[ismiss]
    # Step 2(c)
    mss <- mean(((Xna - Xapp)[!ismiss])^2)
    rel_err <- (mssold - mss) / mss0
    mssold <- mss
    cat("Iter:", iter, "MSS:", mss,
      "Rel. Err:", rel_err, "\n")
    }
```

- Algorithm converges after ~8 iterations  
- Relative error falls below threshold (`1e-7`)  
- Final mean squared error ≈ 0.369 (on non-missing entries)  

- Evaluate performance:
  - Compute correlation between imputed values and true values

```{r}
#| label: chunk27
#| echo: true
cor(Xapp[ismiss], X[ismiss])
```

- Algorithm 12.1 implemented manually for learning purposes  
- For real applications, use `softImpute` package (CRAN)  
- Provides efficient and scalable matrix completion methods
