---
title: "Lab 4"
subtitle: "Decision Trees, Bagging, Random Forests, and Boosting"
author: "[Aldo Solari]{.orange}"
institute: "_Ca' Foscari University of Venice_"
page-layout: full
execute:
  cache: false
filters:
  - remove-pause.lua
format:
  revealjs:
    auto-stretch: true
    center: true
    html-math-method: katex
    transition: none
    output-file: un_lab4_slides.html
    slide-number: true
    callout-appearance: minimal
    code-line-numbers: true
    theme: [default, ../template.css]
    embed-resources: false
    echo: false
    fig-dpi: 200
    logo: img/cf_logo.png
    footer: "[ISL](https://aldosolari.github.io/ISL)"
    highlight-style: github
  html:
    html-math-method: katex
    echo: false
    callout-appearance: minimal
    theme: [simplex, ../template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: true
    smooth-scroll: true
    code-fold: false
    code-summary: "Show the code"
    fig-dpi: 150
    highlight-style: github
editor:
  markdown:
    wrap: 72
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false
knitr::opts_chunk$set(error = TRUE)
```

# Decision Trees

## Fitting Classification Trees

The `tree` library is used to construct regression
trees.

```{r}
#| label: chunk1
#| echo: true
library(tree)
```


Here we fit a regression tree to the `Boston` data set. First, we create
a training set and fit the tree to the training data.

```{r}
#| label: chunk2
#| echo: true
library(MASS)
set.seed(1)
train <- sample(1:nrow(Boston), nrow(Boston) / 2)
tree.boston <- tree(medv ~ ., Boston, subset = train)
summary(tree.boston)
```

Notice that the output of `summary()` indicates that only four of the
variables have been used in constructing the tree. In the context of a
regression tree, the deviance is simply the sum of squared errors for
the tree.

```{r}
#| label: chunk3
#| echo: true
plot(tree.boston)
text(tree.boston, pretty = 0)
```

The variable `lstat` measures the percentage of individuals with lower
socioeconomic status, while the variable `rm` corresponds to the average
number of rooms. The tree indicates that larger values of `rm`, or lower
values of `lstat`, correspond to more expensive houses. For example, the
tree predicts a median house price of $45{,}400$ for homes in census
tracts in which `rm >= 7.553`.

It is worth noting that we could have fit a much bigger tree by passing
`control = tree.control(nobs = length(train), mindev = 0)` into the
`tree()` function.

Now we use the `cv.tree()` function to see whether pruning the tree will
improve performance.

```{r}
#| label: chunk4
#| echo: true
cv.boston <- cv.tree(tree.boston)
plot(cv.boston$size, cv.boston$dev, type = "b")
```

In this case, the most complex tree under consideration is selected by
cross-validation. However, if we wish to prune the tree, we could do so
as follows.

```{r}
#| label: chunk5
#| echo: true
prune.boston <- prune.tree(tree.boston, best = 5)
plot(prune.boston)
text(prune.boston, pretty = 0)
```

In keeping with the cross-validation results, we use the unpruned tree
to make predictions on the test set.

```{r}
#| label: chunk6
#| echo: true
yhat <- predict(tree.boston, newdata = Boston[-train, ])
boston.test <- Boston[-train, "medv"]
plot(yhat, boston.test)
abline(0, 1)
mean((yhat - boston.test)^2)
```

The test set MSE associated with the regression tree is $35.29$. The
square root of the MSE is therefore around $5.941$, indicating that this
model leads to test predictions that are, on average, within
approximately $5{,}941$ of the true median home value for the census
tract.

## Bagging and Random Forests

Here we apply bagging and random forests to the `Boston` data, using the
`randomForest` package in `R`. Recall that bagging is simply a special
case of a random forest with $m = p$.

```{r}
#| label: chunk7
#| echo: true
library(randomForest)
set.seed(1)
bag.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 12, importance = TRUE)
bag.boston
```

The argument `mtry = 12` indicates that all 12 predictors should be
considered for each split of the tree, so bagging is being performed.
How well does this bagged model perform on the test set?

```{r}
#| label: chunk8
#| echo: true
yhat.bag <- predict(bag.boston, newdata = Boston[-train, ])
plot(yhat.bag, boston.test)
abline(0, 1)
mean((yhat.bag - boston.test)^2)
```

The test set MSE associated with the bagged regression tree is $23.42$,
about two-thirds of that obtained using an optimally pruned single tree.
We could change the number of trees grown by `randomForest()` using the
`ntree` argument.

```{r}
#| label: chunk9
#| echo: true
bag.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 12, ntree = 25)
yhat.bag <- predict(bag.boston, newdata = Boston[-train, ])
mean((yhat.bag - boston.test)^2)
```

Growing a random forest proceeds in exactly the same way, except that we
use a smaller value of the `mtry` argument. By default,
`randomForest()` uses $p/3$ variables when building a random forest of
regression trees, and $\sqrt{p}$ variables when building a random forest
of classification trees. Here we use `mtry = 6`.

```{r}
#| label: chunk10
#| echo: true
set.seed(1)
rf.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 6, importance = TRUE)
yhat.rf <- predict(rf.boston, newdata = Boston[-train, ])
mean((yhat.rf - boston.test)^2)
```

The test set MSE is $20.07$; this indicates that random forests yielded
an improvement over bagging in this case.

Using the `importance()` function, we can view the importance of each
variable.

```{r}
#| label: chunk11
#| echo: true
importance(rf.boston)
```

Two measures of variable importance are reported. The first is based on
the mean decrease of accuracy in predictions on the out-of-bag samples
when a given variable is permuted. The second is a measure of the total
decrease in node impurity that results from splits over that variable,
averaged over all trees.

```{r}
#| label: chunk12
#| echo: true
varImpPlot(rf.boston)
```

The results indicate that across all of the trees considered in the
random forest, the wealth of the community (`lstat`) and the house size
(`rm`) are by far the two most important variables.

## Boosting

Here we use the `gbm` package, and within it the `gbm()` function, to
fit boosted regression trees to the `Boston` data set. We run `gbm()`
with the option `distribution = "gaussian"` since this is a regression
problem.

```{r}
#| label: chunk13
#| echo: true
library(gbm)
set.seed(1)
boost.boston <- gbm(medv ~ ., data = Boston[train, ],
    distribution = "gaussian", n.trees = 5000,
    interaction.depth = 4)
```

The `summary()` function produces a relative influence plot and also
outputs the relative influence statistics.

```{r}
#| label: chunk14
#| echo: true
summary(boost.boston)
```

We see that `lstat` and `rm` are by far the most important variables. We
can also produce partial dependence plots for these two variables.

```{r}
#| label: chunk15
#| echo: true
plot(boost.boston, i = "rm")
plot(boost.boston, i = "lstat")
```

These plots illustrate the marginal effect of the selected variables on
the response after integrating out the other variables. In this case, as
we might expect, median house prices are increasing with `rm` and
decreasing with `lstat`.

We now use the boosted model to predict `medv` on the test set.

```{r}
#| label: chunk16
#| echo: true
yhat.boost <- predict(boost.boston,
    newdata = Boston[-train, ], n.trees = 5000)
mean((yhat.boost - boston.test)^2)
```

The test MSE obtained is $18.39$: this is superior to the test MSE of
random forests and bagging. We can also perform boosting with a
different value of the shrinkage parameter $\lambda$ in (8.10). Here we
take $\lambda = 0.2$.

```{r}
#| label: chunk17
#| echo: true
boost.boston <- gbm(medv ~ ., data = Boston[train, ],
    distribution = "gaussian", n.trees = 5000,
    interaction.depth = 4, shrinkage = 0.2, verbose = FALSE)
yhat.boost <- predict(boost.boston,
    newdata = Boston[-train, ], n.trees = 5000)
mean((yhat.boost - boston.test)^2)
```

In this case, using $\lambda = 0.2$ leads to a lower test MSE than
$\lambda = 0.001$.
