---
title: "Lab 4"
subtitle: "Decision Trees, Bagging, Random Forests, and Boosting"
author: "[Aldo Solari]{.orange}"
institute: "_Ca' Foscari University of Venice_"
page-layout: full
execute:
  cache: false
filters:
  - remove-pause.lua
format:
  revealjs:
    auto-stretch: true
    center: true
    html-math-method: katex
    transition: none
    output-file: un_lab4_slides.html
    slide-number: true
    callout-appearance: minimal
    code-line-numbers: true
    theme: [default, ../template.css]
    embed-resources: false
    echo: false
    fig-dpi: 200
    logo: img/cf_logo.png
    footer: "[ISL](https://aldosolari.github.io/ISL)"
    highlight-style: github
  html:
    html-math-method: katex
    echo: false
    callout-appearance: minimal
    theme: [simplex, ../template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: true
    smooth-scroll: true
    code-fold: false
    code-summary: "Show the code"
    fig-dpi: 150
    highlight-style: github
editor:
  markdown:
    wrap: 72
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false
knitr::opts_chunk$set(error = TRUE)
```

# Decision Trees

## Fitting Classification Trees

- The `tree` library in R is used to construct regression trees

```{r}
#| label: chunk1
#| echo: true
library(tree)
```


- Fit a regression tree using the `Boston` dataset  
- Split data into a training set  
- Train the tree model on the training data

```{r}
#| label: chunk2
#| echo: true
library(MASS)
set.seed(1)
train <- sample(1:nrow(Boston), nrow(Boston) / 2)
tree.boston <- tree(medv ~ ., Boston, subset = train)
summary(tree.boston)
```

- `summary()` shows only 4 variables are used in the tree  
- In regression trees, deviance = sum of squared errors (SSE)

```{r}
#| label: chunk3
#| echo: true
plot(tree.boston)
text(tree.boston, pretty = 0)
```

- `lstat`: % of lower socioeconomic status  
- `rm`: average number of rooms  

- Interpretation:
  - Higher `rm` → higher house prices  
  - Lower `lstat` → higher house prices  
  - Example: `rm ≥ 7.553` → predicted price ≈ $45,400  

- Model flexibility:
  - Larger tree can be grown using `tree.control(..., mindev = 0)`  

- Next step:
  - Use `cv.tree()` to evaluate pruning via cross-validation

```{r}
#| label: chunk4
#| echo: true
cv.boston <- cv.tree(tree.boston)
plot(cv.boston$size, cv.boston$dev, type = "b")
```

- Cross-validation selects the most complex tree in this case  
- Tree can still be pruned manually if desired

```{r}
#| label: chunk5
#| echo: true
prune.boston <- prune.tree(tree.boston, best = 5)
plot(prune.boston)
text(prune.boston, pretty = 0)
```

- Based on cross-validation, the unpruned tree is selected  
- Use the unpruned tree to make predictions on the test set

```{r}
#| label: chunk6
#| echo: true
yhat <- predict(tree.boston, newdata = Boston[-train, ])
boston.test <- Boston[-train, "medv"]
plot(yhat, boston.test)
abline(0, 1)
mean((yhat - boston.test)^2)
```

- Test set MSE ≈ 35.29  
- RMSE ≈ 5.941  
- Predictions are on average within ~$5,941 of true median house values

## Bagging and Random Forests

- Apply bagging and random forests using the `randomForest` package in R  
- Bagging = special case of random forest with $m = p$ (all predictors used at each split)

```{r}
#| label: chunk7
#| echo: true
library(randomForest)
set.seed(1)
bag.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 12, importance = TRUE)
bag.boston
```

- `mtry = 12` → all 12 predictors used at each split  
- This corresponds to bagging  
- Evaluate model performance on the test set

```{r}
#| label: chunk8
#| echo: true
yhat.bag <- predict(bag.boston, newdata = Boston[-train, ])
plot(yhat.bag, boston.test)
abline(0, 1)
mean((yhat.bag - boston.test)^2)
```

- Bagged model test MSE ≈ 23.42  
- Significantly lower than single tree (~2/3 of its error)  
- Number of trees can be controlled with the `ntree` parameter

```{r}
#| label: chunk9
#| echo: true
bag.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 12, ntree = 25)
yhat.bag <- predict(bag.boston, newdata = Boston[-train, ])
mean((yhat.bag - boston.test)^2)
```

- Random forests are built like bagging but with smaller `mtry`  
- Default `mtry`:
  - Regression: $p/3$ 
- Here, `mtry = 6` is used

```{r}
#| label: chunk10
#| echo: true
set.seed(1)
rf.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 6, importance = TRUE)
yhat.rf <- predict(rf.boston, newdata = Boston[-train, ])
mean((yhat.rf - boston.test)^2)
```

- Random forest test MSE ≈ 20.07  
- Improves over bagging performance  
- Use `importance()` to evaluate variable importance

```{r}
#| label: chunk11
#| echo: true
importance(rf.boston)
```

- Two measures of variable importance:
  - Mean decrease in accuracy (via permutation on out-of-bag samples)
  - Mean decrease in node impurity (averaged over all trees)

```{r}
#| label: chunk12
#| echo: true
varImpPlot(rf.boston)
```

- Most important variables in the random forest:
  - `lstat` (wealth / socioeconomic status)  
  - `rm` (house size / number of rooms)

## Boosting

- Use the `gbm` package to fit boosted regression trees  
- Apply `gbm()` to the `Boston` dataset  
- Set `distribution = "gaussian"` for regression problems

```{r}
#| label: chunk13
#| echo: true
library(gbm)
set.seed(1)
boost.boston <- gbm(medv ~ ., data = Boston[train, ],
    distribution = "gaussian", n.trees = 5000,
    interaction.depth = 4)
```

- `summary()` provides:
  - Relative influence plot  
  - Numerical measures of variable importance

```{r}
#| label: chunk14
#| echo: true
summary(boost.boston)
```

- `lstat` and `rm` are the most important variables  
- Partial dependence plots can be generated for these variables

```{r}
#| label: chunk15
#| echo: true
plot(boost.boston, i = "rm")
plot(boost.boston, i = "lstat")
```

- Partial dependence plots show marginal effect of variables on response  
- Effects observed:
  - Higher `rm` → higher house prices  
  - Higher `lstat` → lower house prices  

- Use boosted model to predict `medv` on the test set

```{r}
#| label: chunk16
#| echo: true
yhat.boost <- predict(boost.boston,
    newdata = Boston[-train, ], n.trees = 5000)
mean((yhat.boost - boston.test)^2)
```

- Boosting test MSE ≈ 18.39  
- Outperforms random forests and bagging  
- Model can be tuned via shrinkage parameter $\lambda$  
- Example: use $\lambda = 0.2$

```{r}
#| label: chunk17
#| echo: true
boost.boston <- gbm(medv ~ ., data = Boston[train, ],
    distribution = "gaussian", n.trees = 5000,
    interaction.depth = 4, shrinkage = 0.2, verbose = FALSE)
yhat.boost <- predict(boost.boston,
    newdata = Boston[-train, ], n.trees = 5000)
mean((yhat.boost - boston.test)^2)
```

- Using $\lambda = 0.2$ results in lower test MSE than $\lambda = 0.001$
