---
title: "Lab 2"
subtitle: "Introduction to Statistical Learning - PISE"
author: "[Aldo Solari]{.orange}"
institute: "_Ca' Foscari University of Venice_"
page-layout: full
execute:
  cache: false
filters: 
  - remove-pause.lua
format:
  revealjs:
    auto-stretch: true
    center: true
    html-math-method: katex
    transition: none
    output-file: un_lab2_slides.html
    slide-number: true
    callout-appearance: minimal
    code-line-numbers: true
    theme: [default, ../template.css] # alternative themes (subset): default, night, dark
    embed-resources: false
    echo: false
    fig-dpi: 200
    # incremental: true  # Remove comment if you like incremental bullet points
    logo: img/cf_logo.png
    footer: "[ISL](https://aldosolari.github.io/ISL)"
    highlight-style: github
  html:
    html-math-method: katex
    echo: false
    callout-appearance: minimal
    theme: [simplex, ../template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: true
    smooth-scroll: true
    code-fold: false
    code-summary: "Show the code"
    fig-dpi: 150
    highlight-style: github
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---



# Linear Regression

## Libraries

- `library()` loads *packages*  
- Packages contain functions and datasets not always included in base `R`  
- Basic linear regression tools are in base `R`  
- More specialized tools require additional packages  
- Here we load:
  - `MASS`
  - `ISLR2`

```{r}
#| label: chunk1
#| echo: true
library(MASS)
library(ISLR2)
```

- If a package is not installed, `library()` returns an error  
- Some packages (e.g. `MASS`) usually come with `R`  
- Others (e.g. `ISLR2`) may need to be installed first  
- Install once with `install.packages("ISLR2")`  
- Then load the package in each new `R` session with `library()`  

## Simple Linear Regression

- `Boston` is a dataset in `ISLR2`  
- Response variable: `medv` (median house value)  
- Predictors include variables such as:
  - `rm`
  - `age`
  - `lstat`

```{r}
#| label: chunk2
#| echo: true
head(Boston)
```

- Use `?Boston` to see the dataset documentation  

- Fit a simple linear regression with `lm()`  
- Syntax: `lm(y ~ x, data = dataset)`  
- Here:
  - response = `medv`
  - predictor = `lstat`

```{r}
#| label: chunk3
#| echo: true
#| error: true
lm.fit <- lm(medv ~ lstat)
```

- This gives an error because `R` does not know where `medv` and `lstat` are stored  
- Fix:
  - specify `data = Boston`, or
  - use `attach(Boston)`

```{r}
#| label: chunk4
#| echo: true
lm.fit <- lm(medv ~ lstat, data = Boston)
attach(Boston)
lm.fit <- lm(medv ~ lstat)
```

- Typing `lm.fit` gives basic model output  
- `summary(lm.fit)` gives more details:
  - coefficients
  - standard errors
  - $p$-values
  - $R^2$
  - $F$-statistic

```{r}
#| label: chunk5
#| echo: true
lm.fit
summary(lm.fit)
```

- Use `names()` to see what is stored in the fitted model object  
- Use extractor functions such as `coef()` to access components safely  

```{r}
#| label: chunk6
#| echo: true
names(lm.fit)
coef(lm.fit)
```

- Use `confint()` to compute confidence intervals for coefficients  

```{r}
#| label: chunk7
#| echo: true
confint(lm.fit)
```

- Use `predict()` to compute:
  - confidence intervals for the mean response
  - prediction intervals for a new observation

```{r}
#| label: chunk8
#| echo: true
predict(lm.fit, data.frame(lstat = (c(5, 10, 15))),
    interval = "confidence")
predict(lm.fit, data.frame(lstat = (c(5, 10, 15))),
    interval = "prediction")
```

- Confidence and prediction intervals are centered at the same fitted value  
- Prediction intervals are wider than confidence intervals  

- Plot the data and add the least-squares regression line  

```{r}
#| label: chunk9
#| echo: true
plot(lstat, medv)
abline(lm.fit)
```

- The plot suggests some possible non-linearity  
- We will revisit this later  

- `abline()` can add any line to a plot  
- `abline(a, b)` draws a line with:
  - intercept = `a`
  - slope = `b`
- Plot options:
  - `lwd` controls line width
  - `col` controls color
  - `pch` controls plotting symbol

```{r}
#| label: chunk10
#| echo: true
plot(lstat, medv)
abline(lm.fit, lwd = 3)
abline(lm.fit, lwd = 3, col = "red")
plot(lstat, medv, col = "red")
plot(lstat, medv, pch = 20)
plot(lstat, medv, pch = "+")
plot(1:20, 1:20, pch = 1:20)
```

## Non-linear Transformations of the Predictors

- `lm()` can include transformed predictors  
- Use `I()` when arithmetic operators would otherwise be interpreted specially in a formula  
- Example: `I(lstat^2)` adds a quadratic term  

```{r}
#| label: chunk11
#| echo: true
lm.fit2 <- lm(medv ~ lstat + I(lstat^2), data = Boston)
summary(lm.fit2)
```

- The quadratic term appears to improve the model  
- Compare the linear and quadratic models with `anova()`  

```{r}
#| label: chunk12
#| echo: true
lm.fit <- lm(medv ~ lstat, data = Boston)
anova(lm.fit, lm.fit2)
```

- Model 1: linear fit with `lstat`  
- Model 2: quadratic fit with `lstat` and `lstat^2`  
- `anova()` tests whether the larger model improves fit significantly  
- Very small $p$-value indicates strong evidence in favor of the quadratic model  

- Check the residual diagnostics for the quadratic model  

```{r}
#| label: chunk13
#| echo: true
par(mfrow = c(2, 2))
plot(lm.fit2)
```

- With the quadratic term included, residual patterns are reduced  

- For higher-order polynomials, `poly()` is more convenient than writing each power manually  
- Example: fifth-order polynomial fit  

```{r}
#| label: chunk14
#| echo: true
lm.fit5 <- lm(medv ~ poly(lstat, 5), data = Boston)
summary(lm.fit5)
```

- Higher-order polynomial terms can improve fit  
- In this example, terms beyond fifth order are not significant  

- By default, `poly()` uses orthogonal polynomials  
- These are not the raw powers of the predictor  
- Fitted values are the same as with raw polynomials, but coefficients differ  
- Use `raw = TRUE` if raw powers are desired  

- We are not restricted to polynomial transformations  
- Example: log transformation  

```{r}
#| label: chunk15
#| echo: true
summary(lm(medv ~ log(rm), data = Boston))
```

# Cross-Validation 

- In this lab, we explore *resampling methods*  
- Focus:
  - validation set approach
  - leave-one-out cross-validation (LOOCV)
  - $k$-fold cross-validation  
- Some commands may take a while to run  

## The Validation Set Approach

- Use the validation set approach to estimate test error  
- Work with the `Auto` data set  
- Set a random seed for reproducibility  
- This ensures the same training/validation split each time  

```{r}
#| label: chunk16
#| echo: true
library(ISLR2)
set.seed(1)
train <- sample(392, 196)
```

- Split the data into two halves  
- Here, `train` contains the indices of the training observations  
- Fit the model only on the training set using `subset = train`  

```{r}
#| label: chunk17
#| echo: true
lm.fit <- lm(mpg ~ horsepower, data = Auto, subset = train)
```

- Use `predict()` to compute fitted values for all observations  
- Compute validation MSE on the observations not in the training set  
- `-train` selects the validation observations  

```{r}
#| label: chunk18
#| echo: true
attach(Auto)
mean((mpg - predict(lm.fit, Auto))[-train]^2)
```

- Estimate test MSE for quadratic and cubic models  
- Use `poly()` to fit polynomial regressions  

```{r}
#| label: chunk19
#| echo: true
lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, 
    subset = train)
mean((mpg - predict(lm.fit2, Auto))[-train]^2)
lm.fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, 
    subset = train)
mean((mpg - predict(lm.fit3, Auto))[-train]^2)
```

- Different training/validation splits give slightly different results  
- Repeat with a different random seed  

```{r}
#| label: chunk20
#| echo: true
set.seed(2)
train <- sample(392, 196)
lm.fit <- lm(mpg ~ horsepower, subset = train)
mean((mpg - predict(lm.fit, Auto))[-train]^2)
lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, 
    subset = train)
mean((mpg - predict(lm.fit2, Auto))[-train]^2)
lm.fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, 
    subset = train)
mean((mpg - predict(lm.fit3, Auto))[-train]^2)
```

- Main result:
  - quadratic model improves over linear
  - little evidence that cubic improves further  

## Leave-One-Out Cross-Validation

- LOOCV can be computed with `cv.glm()`  
- `cv.glm()` works with models fit using `glm()`  
- Without a `family` argument, `glm()` performs linear regression  
- So `glm()` and `lm()` give the same fit here  

```{r}
#| label: chunk21
#| echo: true
glm.fit <- glm(mpg ~ horsepower, data = Auto)
coef(glm.fit)
```

```{r}
#| label: chunk22
#| echo: true
lm.fit <- lm(mpg ~ horsepower, data = Auto)
coef(lm.fit)
```

- Use `glm()` because it works with `cv.glm()`  
- `cv.glm()` is in the `boot` package  

```{r}
#| label: chunk23
#| echo: true
library(boot)
glm.fit <- glm(mpg ~ horsepower, data = Auto)
cv.err <- cv.glm(Auto, glm.fit)
cv.err$delta
```

- `cv.glm()` returns a list  
- `delta` contains the cross-validation estimates  
- For LOOCV, the two values are essentially the same  

- Compute LOOCV error for polynomial models of degree 1 to 10  
- Store the results in a vector  

```{r}
#| label: chunk24
#| echo: true
cv.error <- rep(0, 10)
for (i in 1:10) {
  glm.fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
  cv.error[i] <- cv.glm(Auto, glm.fit)$delta[1]
}
cv.error
```

- Main pattern:
  - large drop from linear to quadratic
  - little improvement beyond degree 2  

## $k$-Fold Cross-Validation

- `cv.glm()` can also perform $k$-fold CV  
- Here we use:
  - $k = 10$
- Again evaluate polynomial models of degree 1 to 10  

```{r}
#| label: chunk25
#| echo: true
set.seed(17)
cv.error.10 <- rep(0, 10)
for (i in 1:10) {
  glm.fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
  cv.error.10[i] <- cv.glm(Auto, glm.fit, K = 10)$delta[1]
}
cv.error.10
```

- $k$-fold CV is faster to compute than LOOCV here  
- Results are similar:
  - quadratic model improves over linear
  - higher-order terms add little  

- With LOOCV, the two values in `delta` are almost identical  
- With $k$-fold CV, they differ slightly:
  - first value = standard $k$-fold CV estimate
  - second value = bias-corrected estimate  
- On this data set, the difference is very small  
