---
title: "Lab 3"
subtitle: "Introduction to Statistical Learning - PISE"
author: "[Aldo Solari]{.orange}"
institute: "_Ca' Foscari University of Venice_"
page-layout: full
execute:
  cache: false
filters: 
  - remove-pause.lua
format:
  revealjs:
    auto-stretch: true
    center: true
    html-math-method: katex
    transition: none
    output-file: un_lab3_slides.html
    slide-number: true
    callout-appearance: minimal
    code-line-numbers: true
    theme: [default, ../template.css] # alternative themes (subset): default, night, dark
    embed-resources: false
    echo: false
    fig-dpi: 200
    # incremental: true  # Remove comment if you like incremental bullet points
    logo: img/cf_logo.png
    footer: "[ISL](https://aldosolari.github.io/ISL)"
    highlight-style: github
  html:
    html-math-method: katex
    echo: false
    callout-appearance: minimal
    theme: [simplex, ../template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: true
    smooth-scroll: true
    code-fold: false
    code-summary: "Show the code"
    fig-dpi: 150
    highlight-style: github
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---



# Linear Regression

## Multiple Linear Regression

- Fit multiple regression models with the `lm()` function  
- Syntax: `lm(y ~ x1 + x2 + x3, data = ...)`  

```{r}
#| label: chunk1
#| echo: true
library(MASS)
lm.fit <- lm(medv ~ lstat + age, data = Boston)
summary(lm.fit)
```

- To use all predictors, write `.` as shorthand  

```{r}
#| label: chunk2
#| echo: true
lm.fit <- lm(medv ~ ., data = Boston)
summary(lm.fit)
```

- Components of the summary object can be extracted by name  
- Examples:
  - `summary(lm.fit)$r.sq`
  - `summary(lm.fit)$sigma`

- `vif()` from the `car` package computes variance inflation factors  
- VIFs help assess multicollinearity  

```{r}
#| label: chunk3
#| echo: true
library(car)
vif(lm.fit)
```

- Exclude one predictor using `-` in the formula  
- Example: remove `age`

```{r}
#| label: chunk4
#| echo: true
lm.fit1 <- lm(medv ~ . - age, data = Boston)
summary(lm.fit1)
```

- Alternatively, update an existing model with `update()`  

```{r}
#| label: chunk5
#| echo: true
lm.fit1 <- update(lm.fit, ~ . - age)
```

## Interaction Terms

- Include interaction terms with `:`  
- Example: `lstat:age`  
- Use `*` as shorthand for:
  - main effects
  - interaction term  

- `lstat * age` is shorthand for:
  - `lstat + age + lstat:age`

```{r}
#| label: chunk6
#| echo: true
summary(lm(medv ~ lstat * age, data = Boston))
```


## Qualitative Predictors

- `Carseats` is another dataset in `ISLR2`  
- Response variable: `Sales`  
- Predictors include both quantitative and qualitative variables  

```{r}
#| label: chunk7
#| echo: true
library(ISLR2)
head(Carseats)
```

- `ShelveLoc` is a qualitative predictor  
- Levels:
  - `Bad`
  - `Medium`
  - `Good`
- `R` automatically creates dummy variables for qualitative predictors  

```{r}
#| label: chunk8
#| echo: true
lm.fit <- lm(Sales ~ . + Income:Advertising + Price:Age, 
    data = Carseats)
summary(lm.fit)
```

- Use `contrasts()` to see the dummy-variable coding used by `R`  

```{r}
#| label: chunk9
#| echo: true
attach(Carseats)
contrasts(ShelveLoc)
```

- Use `?contrasts` for more information  

- Here:
  - `ShelveLocGood = 1` if location is good, 0 otherwise
  - `ShelveLocMedium = 1` if location is medium, 0 otherwise
  - `Bad` is the baseline category
- Positive coefficients indicate higher sales relative to the baseline category  

## Writing Functions

- `R` includes many built-in functions  
- We can also write our own functions when needed  
- Example: create a function `LoadLibraries()`  

- Before defining it, calling the function gives an error  

```{r}
#| label: chunk10
#| echo: true
#| error: true
LoadLibraries
LoadLibraries()
```

- Define the function with `function() { ... }`  
- The braces `{}` contain the commands in the function body  

```{r}
#| label: chunk11
#| echo: true
LoadLibraries <- function() {
 library(ISLR2)
 library(MASS)
 print("The libraries have been loaded.")
}
```

- Typing the function name prints its definition  

```{r}
#| label: chunk12
#| echo: true
LoadLibraries
```

- Calling the function executes the commands  

```{r}
#| label: chunk13
#| echo: true
LoadLibraries()
```


## Best Subset Selection

- Apply best subset selection to the `Hitters` data  
- Goal: predict a player's `Salary` using performance statistics from the previous year  

- `Salary` has missing values  
- `is.na()` identifies missing entries  
- `sum()` counts how many are missing  

```{r}
#| label: chunk14
#| echo: true
library(ISLR2)
names(Hitters)
dim(Hitters)
sum(is.na(Hitters$Salary))
```

- `Salary` is missing for 59 players  
- `na.omit()` removes rows with missing values  

```{r}
#| label: chunk15
#| echo: true
Hitters <- na.omit(Hitters)
dim(Hitters)
sum(is.na(Hitters))
```

- `regsubsets()` from the `leaps` package performs best subset selection  
- It finds the best model of each size, where *best* means lowest RSS  
- `summary()` shows the selected variables for each model size  

```{r}
#| label: chunk16
#| echo: true
library(leaps)
regfit.full <- regsubsets(Salary ~ ., Hitters)
summary(regfit.full)
```

- An asterisk indicates that a variable is included in the model  
- By default, output is shown only up to 8 predictors  
- Use `nvmax` to increase the maximum model size  

```{r}
#| label: chunk17
#| echo: true
regfit.full <- regsubsets(Salary ~ ., data = Hitters,
    nvmax = 19)
reg.summary <- summary(regfit.full)
```

- `summary()` also returns:
  - `R^2`
  - RSS
  - adjusted `R^2`
  - `C_p`
  - BIC

```{r}
#| label: chunk18
#| echo: true
names(reg.summary)
```

- `R^2` increases as more variables are added  

```{r}
#| label: chunk19
#| echo: true
reg.summary$rsq
```

- Plot RSS and adjusted `R^2` across model sizes  
- `type = "l"` connects the points with lines  

```{r}
#| label: chunk20
#| echo: true
par(mfrow = c(2, 2))
plot(reg.summary$rss, xlab = "Number of Variables",
    ylab = "RSS", type = "l")
plot(reg.summary$adjr2, xlab = "Number of Variables",
    ylab = "Adjusted RSq", type = "l")
```

- `points()` adds points to an existing plot  
- `which.max()` identifies the location of the maximum value  
- Mark the model with the largest adjusted `R^2`  

```{r}
#| label: chunk21
#| echo: true
which.max(reg.summary$adjr2)
plot(reg.summary$adjr2, xlab = "Number of Variables",
    ylab = "Adjusted RSq", type = "l")
points(11, reg.summary$adjr2[11], col = "red", cex = 2, 
    pch = 20)
```

- Use `which.min()` to identify the best model by `C_p` and BIC  

```{r}
#| label: chunk22
#| echo: true
plot(reg.summary$cp, xlab = "Number of Variables",
    ylab = "Cp", type = "l")
which.min(reg.summary$cp)
points(10, reg.summary$cp[10], col = "red", cex = 2,
    pch = 20)
which.min(reg.summary$bic)
plot(reg.summary$bic, xlab = "Number of Variables",
    ylab = "BIC", type = "l")
points(6, reg.summary$bic[6], col = "red", cex = 2,
    pch = 20)
```

- `plot.regsubsets()` displays selected variables by model size  
- Models can be ranked by:
  - `r2`
  - `adjr2`
  - `Cp`
  - `bic`

```{r}
#| label: chunk23
#| echo: true
plot(regfit.full, scale = "r2")
plot(regfit.full, scale = "adjr2")
plot(regfit.full, scale = "Cp")
plot(regfit.full, scale = "bic")
```

- Use `coef()` to inspect coefficients for a chosen model  
- Example: the 6-variable model  

```{r}
#| label: chunk24
#| echo: true
coef(regfit.full, 6)
```

## Forward and Backward Stepwise Selection

- `regsubsets()` can also perform:
  - forward stepwise selection
  - backward stepwise selection  
- Use `method = "forward"` or `method = "backward"`  

```{r}
#| label: chunk25
#| echo: true
regfit.fwd <- regsubsets(Salary ~ ., data = Hitters,
    nvmax = 19, method = "forward")
summary(regfit.fwd)
regfit.bwd <- regsubsets(Salary ~ ., data = Hitters,
    nvmax = 19, method = "backward")
summary(regfit.bwd)
```

- Best subset, forward, and backward selection may agree for small models  
- They can differ for larger models  

```{r}
#| label: chunk26
#| echo: true
coef(regfit.full, 7)
coef(regfit.fwd, 7)
coef(regfit.bwd, 7)
```

## Choosing Among Models Using the Validation-Set Approach and Cross-Validation

- Model size can also be selected using:
  - validation set
  - cross-validation  

- Important: all model-fitting steps must use only the training data  
- Variable selection must also be performed within the training set  

- Create a training/test split  
- `train` is `TRUE` for training observations  
- `test` is `TRUE` for test observations  

```{r}
#| label: chunk27
#| echo: true
set.seed(1)
train <- sample(c(TRUE, FALSE), nrow(Hitters),
    replace = TRUE)
test <- (!train)
```

- Perform best subset selection on the training data only  

```{r}
#| label: chunk28
#| echo: true
regfit.best <- regsubsets(Salary ~ .,
    data = Hitters[train, ], nvmax = 19)
```

- Build a model matrix for the test data  

```{r}
#| label: chunk29
#| echo: true
test.mat <- model.matrix(Salary ~ ., data = Hitters[test, ])
```

- Compute validation error for each model size  
- For each `i`:
  - extract coefficients
  - compute predictions
  - compute test MSE  

```{r}
#| label: chunk30
#| echo: true
val.errors <- rep(NA, 19)
for (i in 1:19) {
 coefi <- coef(regfit.best, id = i)
 pred <- test.mat[, names(coefi)] %*% coefi
 val.errors[i] <- mean((Hitters$Salary[test] - pred)^2)
}
```

- Select the model with the smallest validation error  

```{r}
#| label: chunk31
#| echo: true
val.errors
which.min(val.errors)
coef(regfit.best, 7)
```

- `regsubsets()` has no built-in `predict()` method  
- Define a custom prediction function for reuse  

```{r}
#| label: chunk32
#| echo: true
predict.regsubsets <- function(object, newdata, id, ...) {
  form <- as.formula(object$call[[2]])
  mat <- model.matrix(form, newdata)
  coefi <- coef(object, id = id)
  xvars <- names(coefi)
  mat[, xvars] %*% coefi
}
```

- Refit best subset selection on the full dataset  
- Then extract the chosen model size from the full data  

```{r}
#| label: chunk33
#| echo: true
regfit.best <- regsubsets(Salary ~ ., data = Hitters,
    nvmax = 19)
coef(regfit.best, 7)
```

- The best 7-variable model on the full data may differ from the one on the training set  

- Now use cross-validation for model selection  
- Create:
  - 10 folds
  - a matrix to store CV errors  

```{r}
#| label: chunk34
#| echo: true
k <- 10
n <- nrow(Hitters)
set.seed(1)
folds <- sample(rep(1:k, length = n))
cv.errors <- matrix(NA, k, 19,
    dimnames = list(NULL, paste(1:19)))
```

- For each fold:
  - fit best subset selection on training folds
  - predict on the validation fold
  - store the MSE for each model size  

```{r}
#| label: chunk35
#| echo: true
for (j in 1:k) {
  best.fit <- regsubsets(Salary ~ .,
       data = Hitters[folds != j, ],
       nvmax = 19)
  for (i in 1:19) {
    pred <- predict(best.fit, Hitters[folds == j, ], id = i)
    cv.errors[j, i] <-
         mean((Hitters$Salary[folds == j] - pred)^2)
   }
 }
```

- Average CV errors across folds using `apply()`  

```{r}
#| label: chunk36
#| echo: true
mean.cv.errors <- apply(cv.errors, 2, mean)
mean.cv.errors
par(mfrow = c(1, 1))
plot(mean.cv.errors, type = "b")
```

- Cross-validation selects the model size with minimum average CV error  
- Refit on the full data and extract that model  

```{r}
#| label: chunk37
#| echo: true
reg.best <- regsubsets(Salary ~ ., data = Hitters,
    nvmax = 19)
coef(reg.best, 10)
```
