---
title: "Exercises Unit A - Applied"
subtitle: "Bachelor’s Degree Programme in Philosophy, International and Economic Studies, Ca' Foscari University of Venice."
author: "Aldo Solari"
institute: "Department of Economics, Ca' Foscari University of Venice"
lang: en
page-layout: full
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: true
    smooth-scroll: true
    fig-dpi: 250
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---


### [Homepage](../index.html)

#### Chapter 2 Exercise 8.

This exercise relates to the `College` data set, which can be found in the file `College.csv`. It contains a number of variables for 777 different universities and colleges in the US. The variables are

- `Private` : Public/private indicator  
- `Apps` : Number of applications received  
- `Accept` : Number of applicants accepted  
- `Enroll` : Number of new students enrolled  
- `Top10perc` : New students from top 10% of high school class  
- `Top25perc` : New students from top 25% of high school class  
- `F.Undergrad` : Number of full-time undergraduates  
- `P.Undergrad` : Number of part-time undergraduates  
- `Outstate` : Out-of-state tuition  
- `Room.Board` : Room and board costs  
- `Books` : Estimated book costs  
- `Personal` : Estimated personal spending  
- `PhD` : Percent of faculty with Ph.D.’s  
- `Terminal` : Percent of faculty with terminal degree  
- `S.F.Ratio` : Student/faculty ratio  
- `perc.alumni` : Percent of alumni who donate  
- `Expend` : Instructional expenditure per student  
- `Grad.Rate` : Graduation rate  


Before reading the data into R, it can be viewed in Excel or a text editor.

(a) Use the `read.csv()` function to read the data into R. Call the loaded data college. Make sure that you have the directory set to the correct location for the data.

(b) Look at the data using the `View()` function. You should notice
that the first column is just the name of each university. We don’t
really want R to treat this as data. However, it may be handy to
have these names for later. Try the following commands:

```{r}
#| echo: true
#| eval: false
rownames (college) <- college[, 1]
View(college)
```

You should see that there is now a `row.names` column with the
name of each university recorded. This means that `R` has given
each row a name corresponding to the appropriate university. `R`
will not try to perform calculations on the row names. However,
we still need to eliminate the first column in the data where the
names are stored. Try

```{r}
#| echo: true
#| eval: false
college <- college[, -1]
View(college)
```

Now you should see that the first data column is `Private`. Note
that another column labeled row.names now appears before the
`Private` column. However, this is not a data column but rather
the name that `R` is giving to each row.

(c) i. Use the `summary()` function to produce a numerical summary
of the variables in the data set.

ii. Use the `pairs()` function to produce a scatterplot matrix of
the first ten columns or variables of the data. Recall that
you can reference the first ten columns of a matrix A using
`A[,1:10]`.

iii. Use the `plot()` function to produce side-by-side boxplots of
`Outstate` versus `Private`.

iv. Create a new qualitative variable, called `Elite`, by binning
the `Top10perc` variable. We are going to divide universities
into two groups based on whether or not the proportion
of students coming from the top 10 % of their high school
classes exceeds 50 %.

```{r}
#| echo: true
#| eval: false
Elite <- rep("No", nrow(college))
Elite[college$ Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)
college <- data.frame(college , Elite)
```

Use the `summary()` function to see how many elite universities there are. Now use the `plot()` function to produce
side-by-side boxplots of `Outstate` versus `Elite`.

v. Use the `hist()` function to produce some histograms with
differing numbers of bins for a few of the quantitative variables. You may find the command `par(mfrow = c(2, 2))`
useful: it will divide the print window into four regions so
that four plots can be made simultaneously. Modifying the
arguments to this function will divide the screen in other
ways.

vi. Continue exploring the data, and provide a brief summary
of what you discover.
